---
title: Introduction Statistical Data Simulation
description: Using Statistical Distributions to Generate Data that Mimics the Real World Scenario
author: 
  - name: Cornelius Tanui
    url: https://corneliustanui.rbind.io/
date: '2024-09-12'

format: 
  html: 
    fontfamily: libertinus
    fontsize: 12pt
    # page-layout: article
    html-math-method: katex
    code-fold: true
    link-citations: true
    number-sections: true

slug: classification
categories: [R, simulation, mathematical statistical]
---

```{r chunk options, include = FALSE}

knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)

```

##### Data Simulation

```{r}

# load packages
library(here)
library(tidyverse)

```

Simulation of data used by the Government in MTI to place university students into 5 bands for award of Government scholarship

::: {.callout-important appearance="default" collapse="false"}
## Disclaimer!

The data is **simulated**, and therefore substantially differs with the actual scenario! The data is for learning purposes only, and the statistical estimates reported MUST NOT be taken as true reflection of the real-word situation.
:::

```{r gen data}
# table display setup
#| label: tbl-simulated_data .striped .hover .primary .bordered
#| tbl-cap: "Simulated data"
#| tbl-cap-location: bottom 

# for reproducibility
set.seed(1) 

# generate data
simulated_data <- data.frame(Bands = rbinom(n = 1000, 
                                        size = 5, 
                                        prob = 0.5),
                         Gross_Family_Income = rnbinom(n = 1000, 
                                                         size = 5, 
                                                         mu = 20000),
                         Geographical_Location = as.factor(ceiling(runif(n = 1000, 
                                                                           min = 1, 
                                                                           max = 47))),
                         Poverty_Probability_Index = abs(rnorm(n = 1000, 
                                                                 mean = 0.3, 
                                                                 sd = 0.2)),
                         Orphans = as.factor(rbinom(n = 1000,
                                                    size = 1,
                                                    prob = 0.4)),
                         Disability = as.factor(rbinom(n = 1000,
                                                       size = 1,
                                                       prob = 0.022)),
                         Number_of_Dependents = rpois(n = 1000, 
                                                        lambda = 4),
                         Program_Costs_KES = abs(rnorm(n = 1000,
                                                     mean = 500000,
                                                     sd = 50000)),
                         Gender = as.factor(rbinom(n = 1000,
                                                   size = 3,
                                                   prob = 0.017)))

# view data (printed on your browser)
knitr::kable(head(x = simulated_data, n = 5))

# write data to disc
write.csv(x = simulated_data, row.names = FALSE, file = here::here("./Data/simulated_data.csv"))
saveRDS(object = simulated_data, file = here::here("./Data/simulated_data.rds"))

```
